Predicting complex diseases: performance and robustness [OF NEW METHOD?]

### 2018-03-19 paper/draft.tex
add one ref
add ref dudbridge

In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?]. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part/component/term.. of the phenotypes (a ``simple'' model with only linear effects, and a ``fancy'' model which combines linear, dominant and ..interaction/interaction-type.. effects).
For the ``simple'' model, we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy'' model, we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$). 
We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ 
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. 
Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still ..display/have.. a marginal effect. 
For both models, the ..$y_i$/values of $y_i$.. are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.

We ..make/construct/design/implement/work with.. 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000. It compares all methods described in section \ref{sec:methods} ..on/for/along/in relation to.. all other combinations of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}). Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more than one hour of computation for a single simulation only. 
Scenario \textnumero2 consists of 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of chromosome 6 only. Reducing the number of SNPs aims at increasing the polygenicity of the simulated models and at ..virtually/artificially.. increasing the sample size [REF? DUDBRIDGE?]. For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of the PRS and the ``logit-simple'' methods. 100 simulations are run using 300 causal SNPs anywhere on the genome.

>>>>

In order to simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only chromosome 6), the size of the training set ($\{1000, 2000, \dots, 6000\}$), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part/component/term.. of the phenotypes (a ``simple'' model with only linear effects, and a ``fancy'' model which combines linear, dominant and ..interaction/interaction-type.. effects).
For the ``simple'' model, we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy'' model, we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$). 
We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ 
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. 
Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still ..display/have.. a marginal effect. 
For both models, the ..$y_i$/values of $y_i$.. are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.

We ..make/construct/design/implement/work with.. 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
Scenario \textnumero1 uses the whole dataset (all 22 autosomal chromosomes) and a training set of size 6000. It compares all methods described in section \ref{sec:methods} ..on/for/along/in relation to.. all other combinations of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}). Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more than one hour of computation for a single simulation only. 
Scenario \textnumero2 consists of 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of chromosome 6 only. Reducing the number of SNPs aims at increasing the polygenicity of the simulated models and at ..virtually/artificially.. increasing the sample size \cite[]{dudbridge2013power}. For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
Finally, scenario \textnumero3 reuses the whole dataset but this time varying the size of the training set in order to assess how the sample size affects predictive performance of the PRS and the ``logit-simple'' methods. 100 simulations are run using 300 causal SNPs anywhere on the genome.
